# Tensortree Examples

We've seen that the `tensortree` package can help illuminate tensors via custom printouts, for example in exploring the `mnist` dataset.

```{r}
library(keras)
library(tensortree)

mnist <- dataset_mnist()   # load mnist dataset

train_images <- mnist$train$x
train_labels <- mnist$train$y
```

To start with, we need to add some metadata to the tensors by converting them with `as.tensortree()` or `tt()` for short:

```{r}
print(tt(train_images))
```

Using `%>%` syntax, the above would be `train_images %>% tt() %>% print()`. The `print()` function can take a variety of parameters, for example the number of entries to show for the last 1, 2, or 3 ranks (default = 6).

```{r}
train_images %>% tt() %>% print(end_n = 16)
```

And we can adjust how many entries to show per level above that:

```{r smalltext}
train_images %>% tt() %>% print(end_n = 16, max_per_level = 1)
```

Let's play with some other data by using `keras`' `image_load()` and `image_to_array()` function to read two JPG files into tensors.

[Strawberry](images/strawberry.jpg) [Apple](images/apple.jpg)

For ease of visualization we're going to scale them to just 10x10, and for now we'll format them with `channels_first`, resulting in rank-3 tensors with shape `(3, 10, 10)`.

```{r}
# read raw image data,
# convert to tensor with a 'channels-first' representation
# and convert to tt() 
strawberry <- image_load("images/strawberry.jpg", target_size = c(10, 10)) %>%
              image_to_array(data_format = "channels_first") %>% 
              tt()

apple <- image_load("images/apple.jpg", target_size = c(10, 10)) %>%
         image_to_array(data_format = "channels_first") %>%
         tt()


apple %>% print(max_per_level = 3)
```

We can 'bind' tensors together to create a new tensor of higher rank; in this case a `(3, 10, 10)` tensor bound to a `(3, 10, 10)`
tensor will result in a `(2, 3, 10, 10)` tensor (images, channels, rows, columns).

```{r}
images <- bind(apple, strawberry)

images
```

We can permute the ranks of a tensortree if we wish, for example to convert our channels-first tensor into a channels-last tensor (shape `(2, 10, 10, 3)`).^[This is a fairly thin wrapper around the base-R function `aperm()`.]

```{r}
images <- images %>% permute(c(1, 3, 4, 2))
```

If the last rank of a tensor is of size 3, the printout assumes the tensor represents a color channels-last image, and so prints the last three ranks in "3d", where the channel values are shown as embedded vectors:

```{r}
images %>% print(end_n = 5)
```


